---
title: "Case-study"
author: "Jayanth"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# Process phase:
## install packages:

install the packages:

```{r install packages}

# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("geosphere")
# install.packages("ggplot2")
# install.packages("dplyr")

```

## import libraries:

```{r loading libs} 
library(tidyverse)
library(lubridate)
library(geosphere)
library(ggplot2)
library(dplyr)

```


## loading the dataset:

```{r loading dataset}

jan <- read.csv("dataset/202201-divvy-tripdata.csv")
feb <- read.csv("dataset/202202-divvy-tripdata.csv")
mar <- read.csv("dataset/202203-divvy-tripdata.csv")
apr <- read.csv("dataset/202204-divvy-tripdata.csv")
may <- read.csv("dataset/202205-divvy-tripdata.csv")
jun <- read.csv("dataset/202206-divvy-tripdata.csv")
jul <- read.csv("dataset/202207-divvy-tripdata.csv")
aug <- read.csv("dataset/202208-divvy-tripdata.csv")
sep <- read.csv("dataset/202209-divvy-tripdata.csv")
oct <- read.csv("dataset/202210-divvy-tripdata.csv")
nov <- read.csv("dataset/202211-divvy-tripdata.csv")
dec <- read.csv("dataset/202212-divvy-tripdata.csv")

```

## merging all the data to create year data:


```{r creatig year_data}
year_data <- bind_rows(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec)

```

## creating new columns day and month:

```{r creating weekday and month}
year_data$started_at <- as.POSIXct(year_data$started_at, format = "%Y-%m-%d %H:%M:%S")
year_data$ended_at <- strptime(year_data$ended_at, format = "%Y-%m-%d %H:%M:%S")

year_data <- year_data %>% 
  mutate(ride_weekday = wday(started_at, label = TRUE, abbr = FALSE),
         ride_month = month(started_at, label = TRUE, abbr = FALSE))
```



## add time in hours form start to end

```{r creating time column}
year_data <- year_data %>% 
  mutate(ride_time_mins = as.numeric(difftime(ended_at, started_at, units="mins")))
```



## ride_distance:

```{r adding ride distance}
year_data$ride_distance <- distHaversine(matrix(c(year_data$start_lng, year_data$start_lat)
                                                , ncol = 2),
                 matrix(c(year_data$end_lng, year_data$end_lat), ncol = 2))

year_data$ride_distance <- year_data$ride_distance/ 1000
```


## basic cleaing:
### 1. removing nulls

```{r dropping na}
year_data <- drop_na(year_data)
```


### 2. removing negative distances 

```{r removing neg distances}
year_data <- year_data %>%
  filter(ride_distance>0)

```


# Analyze phase:

## finding no.of casual and member rides took place in the last year.

```{r no_of casual and members}
per_membership <- year_data %>% summarize(
  casual_percentage = sum(member_casual=="casual")*100/n(),
  member_percentage = sum(member_casual=="member")*100/n()
)
  
per_membership
```



## finding different rideable_type took place in the last year.

```{r rideable types}
rideable_types <- unique(year_data$rideable_type)
print(rideable_types)

per_ridetypes <- year_data %>% summarize(
  electric_bike_percentage = sum(rideable_type=="electric_bike")*100/n(),
  classic_bike_percentage = sum(rideable_type=="classic_bike")*100/n(),
  docked_bike_percentage = sum(rideable_type=="docked_bike")*100/n()
)

per_ridetypes
```


## summarize rideable types and group by their membership:

```{r rideable types groupby membership}
per_ridetypes_group_menbership <- year_data %>% group_by(member_casual) %>% summarize(
  electric_bike_percentage = sum(rideable_type=="electric_bike")*100/n(),
  classic_bike_percentage = sum(rideable_type=="classic_bike")*100/n(),
  docked_bike_percentage = sum(rideable_type=="docked_bike")*100/n()
)

per_ridetypes_group_menbership
```


## monthly and weekly rides:

### monthly:
```{r monthly summarize data}

monthly_summarize_data <- year_data %>% summarise(
  Jan = sum(ride_month=="January")*100/n(),
  Feb = sum(ride_month=="February")*100/n(),
  Mar = sum(ride_month=="March")*100/n(),
  Apr = sum(ride_month=="April")*100/n(),
  May = sum(ride_month=="May")*100/n(),
  Jun = sum(ride_month=="June")*100/n(),
  Jul = sum(ride_month=="July")*100/n(),
  Aug = sum(ride_month=="August")*100/n(),
  Sep = sum(ride_month=="September")*100/n(),
  Oct = sum(ride_month=="October")*100/n(),
  Nov = sum(ride_month=="November")*100/n(),
  Dec = sum(ride_month=="December")*100/n(),
)

monthly_summarize_data
```


### monthly_summarize data group by membership:

```{r monthly_summarize_groupby_member_data}
monthly_summarize_groupby_member_data <- year_data %>% group_by(member_casual) %>% summarise(
  Jan = sum(ride_month=="January")*100/n(),
  Feb = sum(ride_month=="February")*100/n(),
  Mar = sum(ride_month=="March")*100/n(),
  Apr = sum(ride_month=="April")*100/n(),
  May = sum(ride_month=="May")*100/n(),
  Jun = sum(ride_month=="June")*100/n(),
  Jul = sum(ride_month=="July")*100/n(),
  Aug = sum(ride_month=="August")*100/n(),
  Sep = sum(ride_month=="September")*100/n(),
  Oct = sum(ride_month=="October")*100/n(),
  Nov = sum(ride_month=="November")*100/n(),
  Dec = sum(ride_month=="December")*100/n(),
)

monthly_summarize_groupby_member_data
```


### Week Day:

```{r weekly_summarize_data}
weekly_summarize_data <- year_data %>% summarise(
  Mon = sum(ride_weekday=="Monday")*100/n(),
  Tue = sum(ride_weekday=="Tuesday")*100/n(),
  Wed = sum(ride_weekday=="Wednesday")*100/n(),
  Thu = sum(ride_weekday=="Thursday")*100/n(),
  Fri = sum(ride_weekday=="Friday")*100/n(),
  Sat = sum(ride_weekday=="Saturday")*100/n(),
  Sun = sum(ride_weekday=="Sunday")*100/n(),
)

weekly_summarize_data
```


### week_summarize data group by membership:

```{r weekly_summarize_groupby_member_data}
weekly_summarize_groupby_member_data <- year_data %>% group_by(member_casual) %>% summarise(
  Mon = sum(ride_weekday=="Monday")*100/n(),
  Tue = sum(ride_weekday=="Tuesday")*100/n(),
  Wed = sum(ride_weekday=="Wednesday")*100/n(),
  Thu = sum(ride_weekday=="Thursday")*100/n(),
  Fri = sum(ride_weekday=="Friday")*100/n(),
  Sat = sum(ride_weekday=="Saturday")*100/n(),
  Sun = sum(ride_weekday=="Sunday")*100/n(),
)

weekly_summarize_groupby_member_data
```


## average ride times

```{r average ride time}
avg_ridetime <- year_data %>% summarise(
  avg_ride_time_in_mins = mean(ride_time_mins)
)

avg_ridetime
```


### group by membership and ride types:

```{r average ride time group by membership}
avg_ridetime_group <- year_data %>% group_by(member_casual, rideable_type) %>% 
  summarise(avg_ride_time_in_mins = mean(ride_time_mins)
)

avg_ridetime_group
```


# Share phase:

## no.of casual and membership raiders:

```{r casual and membership raiders}
ggplot(data = year_data) +
  geom_bar(mapping = aes(x = member_casual, y = ..count..), stat = "count", fill="#8c1c0b") +
  geom_text(mapping = aes(x = member_casual, y = ..count.., label = paste0(round((..count.. / sum(..count..)) * 100), "%")),
    stat = "count",vjust = -0.5,size = 4) +
  labs(title = "No. of Casual and Membership Raiders")
```



## different types of rides used:

```{r differnt ride types}
ggplot(data = year_data) +
  geom_bar(mapping = aes(x=rideable_type, y = ..count..), stat = "count" ,fill = "#91aa31") +
  geom_text(mapping = aes(x=rideable_type, y = ..count.., label = paste0(round((..count.. / sum(..count..)) * 100), "%")),
            stat = "count",vjust = -0.5,size = 4)+
  labs(title = "Different ride types")
```


## different rides and membership:

```{r different rides and membership}
ggplot(data = year_data) +
  geom_bar(mapping = aes(x = member_casual, fill = rideable_type))+
  labs(title = "different ride types and membership")

```


## Monthly bike rides:

```{r monthly bike rides}
ggplot(data = year_data) +
  geom_bar(mapping = aes(x=ride_month, y = ..count.., fill = after_stat(count))
           , stat = "count") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Monthly Bike rides")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Monthly bike rides and their membership

```{r Monthly bike rides and their membership}
ggplot(data = year_data) +
  geom_bar(mapping = aes(x=ride_month, fill=member_casual),
           stat = "count") +
  labs(title = "Monthly Bike rides and membership")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## weekly bike rides and membership

```{r weekly bike rides and membership}
ggplot(data = year_data) +
  geom_bar(mapping = aes(x=ride_weekday, y = ..count.., fill = after_stat(count)),
           stat = "count") +
  labs(title = "Weekday Bike rides and membership")+
  scale_fill_gradient(low = "#a5cfbf", high = "#207a58") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~member_casual)

```
